Expanding the scope of experimental archaeology using the Perception-Process-Product conceptual framework

### 2023-06-03 manuscript/manuscript_word.Rmd
minor updates.

---
title: "Expanding the scope of experimental archaeology using the Perception-Process-Product conceptual framework"
author: 
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
abstract: "This paper presents the outline of the Perception-Process-Product ('Triple P') conceptual framework that aims to expand the scope of experimental archaeology. The Triple P framework emphasizes the amplification of multi-level variabilty and the identification of causal relationships of variations across the levels of perception, process, and product. Here we propose the following five basic measures to put the Triple P framework into practice: 1) the acknowledgement of the contribution and limitations of actualistic experiments properly; 2) the normalization the ethological and ethnographic data collection in experimental projects; 3) the involvement of avocational as well as novice participants; 4) the collaboration across labs on a global scale; and 5) the development of an open-access repository for data reuse.  \\par \\textbf{Keywords:} Experimental archaeology; Ethological analysis; Ethnographical analysis; The curse of knowledge; Collaborative knowledge production"
date: "`r Sys.Date()`"
fontsize: 11pt
output: bookdown::pdf_document2
header-includes:
   - \usepackage{textcomp}
   - \usepackage{utopia}
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage{caption}
   - \captionsetup[figure]{font=scriptsize}
   - \captionsetup[table]{font=scriptsize}
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
---

@@

This paper presents the Perception-Process-Product (hereinafter referred to as "Triple P") conceptual framework to expand the scope of experimental archaeology, which tends to center around the reverse engineering of a past technology in a minimal or least-effort manner while ignoring the rich contextual information it affords. Built upon early intellectual principles and practices in behavioral archaeology [@schiffer2010], the Triple P framework aims to **a)** amplify the expression of variability in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) and **b)** better identify the complex interacting relationships across these three levels of variations. To accomplish these two objectives, we advocate the following five measures as integral components of the Triple P framework: 1) acknowledging the contribution and limitations of actualistic experiments properly; 2) normalizing the ethological and ethnographic data collection in experimental projects; 3) encouraging the involvements of avocational as well as novice participants; 4) boosting the collaboration across labs on a global scale; 5) building an open-access repository for data reuse. It is no doubt that strategies of data collection and analysis of a given experimental project should be primarily derived from the research question, which can be legitimately narrow in scope, but the awareness of the rich toolkit available can sometimes inspire researchers to ask questions that are bold and transformative [@schmidt2020]. Here I will mainly leverage the extensive corpus in experimental designs and inferences revolving around stone artifacts to clarify its meaning and demonstrate the necessity and potentials of this framework.

@@

1)  True controlled experiments are rare. Randomized controlled experiments *sensu stricto* are extremely rare in experimental archaeology when human participants were involved. Rather, most of our knowledge regarding the past are derived from datasets that can be characterized as Small, Unbalanced, Noisy, but Genuine (SUNG) [@arnaud2023]. It has also been a debatable issue in philosophy of science whether Randomized Controled Trial (RCT) represents the golden standard of knowledge [@cartwright2007].
2)  Trade-off between causality and generalizability in experimental design. For example, in a series of recently published experimental studies aiming at understanding the role of language in the transmission of lithic technologies, standardized procedures like video teaching or using bricks as raw materials are very often.

@@

As implied in its name, the implementation of Triple P framework involves the collection of process-level (ethological) and perception-level (ethnographic) data (**Figure** \@ref(fig:concept)), which is critical to address equifinality and multifinality [@hiscock2004; @premoEquifinalityExplanationThoughts2010], two daunting challenges in archaeological inference that partially contributed to the discipline-wide paradigm shift in the 1980s [@lake2014: 264-265]. Equifinality refers to the phenomenon where a similar state or consequence can be achieved through multiple different paths, while multifinality emerges when a similar process can lead to multiple ends. While we cannot fully solve these two problems and accurately reconstruct the past behavioral processes and intentionality simply based on materials remains, context-rich experiments involving the collection of ethological and ethnographic data can help us better document an enlarged range of possible combinations of variation at these three levels and thereby evaluate the probability of certain behavioral mechanisms behind a given archaeological assemblage.

```{r concept, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="A schematic diagram demonstrating how to operationalize the Perception-Process-Product conceptual framework."}
knitr::include_graphics(here::here("figure/Fig1.jpg"))
```

>>>>

---
title: "Expanding the scope of experimental archaeology using the Perception-Process-Product conceptual framework"
author: 
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
abstract: "This paper presents the outline of the Perception-Process-Product ('Triple P') conceptual framework that aims to expand the scope of experimental archaeology. The Triple P framework emphasizes the amplification of multi-level variabilty and the identification of causal relationships of variations across the levels of perception, process, and product. Here we propose the following five basic measures to put the Triple P framework into practice: 1) the acknowledgement of the contribution and limitations of actualistic experiments properly; 2) the normalization the ethological and ethnographic data collection in experimental projects; 3) the involvement of avocational as well as novice participants; 4) the collaboration across labs on a global scale; and 5) the development of an open-access repository for data reuse.  \\par \\textbf{Keywords:} Experimental archaeology; Ethological analysis; Ethnographical analysis; The curse of knowledge; Collaborative knowledge production"
date: "`r Sys.Date()`"
output:
  word_document
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
fontsize: 11pt
always_allow_html: true
---

@@

This paper presents the Perception-Process-Product (hereinafter referred to as "Triple P") conceptual framework to expand the scope of experimental archaeology, which tends to center around the reverse engineering of a past technology in a minimal or least-effort manner while ignoring the rich contextual information it affords. Built upon early intellectual principles and practices in behavioral archaeology [@schiffer2010], the Triple P framework aims to a) amplify the expression of variability in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) and b) better identify the complex interacting relationships across these three levels of variations. To accomplish these two objectives, we advocate the following five measures as integral components of the Triple P framework: 1) acknowledging the contribution and limitations of actualistic experiments properly; 2) normalizing the ethological and ethnographic data collection in experimental projects; 3) encouraging the involvements of avocational as well as novice participants; 4) boosting the collaboration across labs on a global scale; 5) building an open-access repository for data reuse. It is no doubt that strategies of data collection and analysis of a given experimental project should be primarily derived from the research question, which can be legitimately narrow in scope, but the awareness of the rich toolkit available can sometimes inspire researchers to ask questions that are bold and transformative [@schmidt2020]. Here I will mainly leverage the extensive corpus in experimental designs and inferences revolving around stone artifacts to clarify its meaning and demonstrate the necessity and potentials of this framework.

@@

True controlled experiments are rare. Randomized controlled experiments sensu stricto are extremely rare in experimental archaeology when human participants were involved. Rather, most of our knowledge regarding the past are derived from datasets that can be characterized as Small, Unbalanced, Noisy, but Genuine (SUNG) [@arnaud2023]. It has also been a debatable issue in philosophy of science whether Randomized Controled Trial (RCT) represents the golden standard of knowledge [@cartwright2007].

Trade-off between causality and generalizability in experimental design. For example, in a series of recently published experimental studies aiming at understanding the role of language in the transmission of lithic technologies, standardized procedures like video teaching or using bricks as raw materials are very often.

@@

As implied in its name, the implementation of Triple P framework involves the collection of process-level (ethological) and perception-level (ethnographic) data (Figure @ref(fig:concept)), which is critical to address equifinality and multifinality [@hiscock2004; @premoEquifinalityExplanationThoughts2010], two daunting challenges in archaeological inference that partially contributed to the discipline-wide paradigm shift in the 1980s [@lake2014: 264-265]. Equifinality refers to the phenomenon where a similar state or consequence can be achieved through multiple different paths, while multifinality emerges when a similar process can lead to multiple ends. While we cannot fully solve these two problems and accurately reconstruct the past behavioral processes and intentionality simply based on materials remains, context-rich experiments involving the collection of ethological and ethnographic data can help us better document an enlarged range of possible combinations of variation at these three levels and thereby evaluate the probability of certain behavioral mechanisms behind a given archaeological assemblage.

{r concept, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="A schematic diagram demonstrating how to operationalize the Perception-Process-Product conceptual framework."}
knitr::include_graphics(here::here("figure/Fig1.jpg"))
